---
title: "ESM 244 Lab 1"
author: "Samantha Kuglen"
date: "1/22/2022"
output: html_document
---

```{r setup, include=TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(here)
library(sf)
library(tmap)

```
### Read in the data
```{r}
sf_trees <- read_csv(here("data", "sf_trees", "sf_trees.csv"),
                     show_col_types = FALSE)

# in console names(sf_trees) to view 

```

### Part 1: wrangling and ggplot review

**Example 1:** Find counts and observation by `legal_status` and wrangle a bit.

```{r}
### method 1: group_by() %>%  summarize()

sf_trees %>% 
  group_by(legal_status) %>% 
  summarize(tree_count = n())

### method 2: different way plus a few new functions
top_5_status <- sf_trees %>% 
  count(legal_status) %>% 
  drop_na(legal_status) %>% 
  rename(tree_count = n) %>% 
  relocate(tree_count) %>% 
  slice_max(tree_count, n = 5) %>% 
  arrange(-tree_count) #desc
```

Make a graph of the top 5 from above:
```{r}
ggplot(data = top_5_status, aes(x = fct_reorder(legal_status, tree_count), y = tree_count)) + 
  geom_col(fill = "darkgreen") +
  labs(x = "legal status", y = "tree count") +
  coord_flip() +
  theme_minimal()
  
```

**Example 2:** Only going to keep observations where legal status is "Permitted Site" and caretaker is "MTA" and store as `permitted_data_df`
shift-cmd-c to comment
```{r}
# sf_trees$legal_status %>%  unique() 
# unique(sf_trees$caretaker)
# & or | or
# %in% operatpor
permitted_data_df <- sf_trees %>% 
  filter(legal_status %in% c("Permitted Site", "Private") & caretaker == "MTA")
```

**Example 3:** Only keep Blackwood Acacia trees, and then only keep columns `legal_status`, `date`, `latitude`, `longitude`, and store as `blackwood_acacia_df`

```{r}
blackwood_acacia_df <- sf_trees %>% 
  filter(str_detect(species, 'Blackwood Acacia')) %>% 
  select(legal_status, date, lat = latitude, lon = longitude)

### Make a little graph of locations
ggplot(data = blackwood_acacia_df, aes(x = lon, y = lat)) +
  geom_point(color = "darkgreen")
```

**Example 4:** use `tidyr::separate()`
```{r}
sf_trees_sep <- sf_trees %>% 
  separate(species, into = c("spp_scientific", "spp_common"), sep = ' :: ')
```

**Example 5:** use `tidyr::unit()`
```{r}
ex_5 <- sf_trees %>% 
  unite('id_status', tree_id, legal_status, sep = '_COOL_')
```

